### 2018-10-20 write-up/paper/lang_and_examples.tex
Add array sum example

However, linearity by itself is not sufficient to produce an expressive enough
programming language. For values such as booleans, integers, floating-point
numbers as well as pure functions, we need to be able to use them
\emph{intuitionistically}, that is, more than once or not at all. For this
reason, we have the !-constructor at the type level and its corresponding
\highl{Many}-constructor and \highl{let Many <id> = .. in ..}-eliminator at
the term level. Because we want to restrict how a programmer can alias pointers
and prevent a programmer from ignoring them (a memory leak), \lang\ enforces
simple syntactic restrictions on which values can be wrapped up in a
\highl{Many} constructor (details in Section~\ref{sec:formal_system}).

@@

Matrix unsharing is one of only \emph{two} operations that can fail at runtime
(the other being dimension checks, such as for \highl{gemm}). The check being
performed is a simple sanity check that the two aliasing pointers passed to
\highl{unshare} point to the same array/matrix.
Section~\ref{sec:discussion_related_work} contains an overview of how we could
remove the need for this by tracking pointer identities statically by
augmenting the type system further.

@@

Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
an aggresively pedestrian first example, in a linearly typed language such as
\lang\ it represents the culmination of a rather number of features.

@@

@@

>>>>

However, linearity by itself is not sufficient to produce an expressive enough
programming language. For values such as booleans, integers, floating-point
numbers as well as pure functions, we need to be able to use them
\emph{intuitionistically}, that is, more than once or not at all. For this
reason, we have the ! constructor at the type level and its corresponding
\highl{Many} constructor and \highl{let Many <id> = .. in ..} eliminator at
the term level. Because we want to restrict how a programmer can alias pointers
and prevent a programmer from ignoring them (a memory leak), \lang\ enforces
simple syntactic restrictions on which values can be wrapped up in a
\highl{Many} constructor (details in Section~\ref{sec:formal_system}).

@@

Aside from out-of-bounds indexing, matrix unsharing is one of only \emph{two}
operations that can fail at runtime (the other being dimension checks, such as
for \highl{gemm}). The check being performed is a simple sanity check that the
two aliasing pointers passed to \highl{unshare} point to the same array/matrix.
Section~\ref{sec:discussion_related_work} contains an overview of how we could
remove the need for this by tracking pointer identities statically by
augmenting the type system further.

@@

Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
an aggresively pedestrian first example, in a linearly typed language such as
\lang\ it represents the culmination of many features.

@@

Now we can add fractional permissions to the mix:
Figure~\ref{fig:lang_sumarray} shows a simple, tail-recursive implementation of
summing all the elements in an array. There are many new features; first among
them is \highl{!x0 : !elt}, the type of array/matrix elements (64-bit floating
point).

@@

Second is \highl{('x) (row : 'x arr)} which is an array with a
universally-quantified fractional permission. In particular, this means the
body of the function cannot mutate or free the input array, only read from it.
If the programmer did try to mutate or free \highl{row}, then they would get a
helpful error message (Figure~\ref{fig:lang_errormsg}).

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    \lang.}\label{fig:lang_errormsg}
\end{figure}

Alongside taking a \highl{row : 'x arr}, the function also returns an array
with exactly the same fractional permission as the \highl{row} (which can only
be \highl{row}).  This is necessary because of linearity: for the caller, the
original array passed in as an argument would be out of scope for the rest of
the expression, so it needs to be returned and then rebound to be used for the
rest of the function.

An example of this consuming and rebinding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive \texttt{get: 'x. 'x arr --o
!int --o 'x arr * !elt}. Although fractional permissions can be passed around
explicitly  (as done in the recursive call), they can also be
\emph{automatically inferred at call sites}, \texttt{row[i] == get \_ row i}
takes advantage of this convenience.
